Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-09 paper/paper.tex
Some edits to section 3

\documentclass{llncs}
% \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\pagestyle{plain}

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion while ensuring termination and providing reasoning about
program equivalences, as well as deriving optimisations which have been
successfully applied to functional programming. Formalising recursion schemes in
a type theory offers additional termination guarantees, but it often requires
compromises affecting the resulting code, such as imposing performance
penalties, requiring the assumption of additional axioms, or introducing unsafe
casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).

@@

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

Despite the rigorous set of algebraic laws that are satisfied by recursion
schemes, there is a lack of tool support for their use in the context of program
calculation. In fact, most of the work on applying program calculation is done
by performing pen-and-paper proofs, and then translating the result to specific
instances of recursion schemes, implemented in a programming language (generally
Haskell). There are some examples of implementations of program calculation
techniques, but these implementations are scarce, not up to date, and not
verified. For example, Cunha et al. automated program calculation techniques by
a custom Haskell implementation~\cite{DrHylo}.

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the work
in mechanising recursion schemes focuses on a narrow subset of the known
recursion schemes due to termination issues, or do not focus on proving
algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.
Indeed, most of the mechanisations of generic
recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
on the \emph{unfolds}. In general, Hinze et al.~\cite{HinzeWG15} has proven that
all known  recursion schemes are instances of a  divide-and-conquer algorithm
which goes under the name of \emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To
date, no work has mechanised hylomorphisms in the Coq proof assistant, together
with their algebraic laws.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration is what allows the execution of code that has been verified in
Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
al.'s approach, extraction (1) does not preserve the recursive structure of
common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
the generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
ANONYMISE THIS}. While programmers still need to reason about the termination of
their programs, our mechanisation will allow the use of the algebraic laws of
hylomorphisms for \emph{program calculation}, as well as extracting to idiomatic
code (see examples of extracted code in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

To summarise, our contributions are as follows. In this paper we provide a
framework for \emph{program calculation} with recursion schemes in Coq, where
program calculation is done by applying regular Coq equalities, based on  proofs
of the algebraic laws of hylomorphisms, and that allows the extraction of safe,
idiomatic OCaml code:
\begin{itemize}
  \item In Section~\ref{sec:containers} we formalise the type of container
functors ensuring the presence of least and greatest fixed-points for functors
and suitably adapted for program extraction
  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
hylomorhisms and their universal properties
  \item In Section~\ref{sec:examples} we use the framework to formalise examples
of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
Furthermore we verify the short-cut fusion optimisation and show the extracted
optimised code to OCaml.
\end{itemize}

\section{Recursion Schemes}
The structure of the a certain data is very similar to the structure of an
algorithm which processes that data. This relationship manifests in the form of
structured recursion schemes which are widely used in functional languages such
Haskell. Canonical examples are folds (catamorphisms), which consume data, and
unfolds (anamorphisms), which produce it. While some implementations of
recursion schemes like $\operator{foldr}$ in Haskell as specific to a particular
data structure, in this case $\operator{List}$s, we can generalise further these
ideas to account for generic (co)inductive data types and generic algorithms
operating on them. This is done by looking at recursion schemes from the point
of view of category theory.

@@

A \emph{functor} $F : \CatC \to \CatD$ is a map between categories mapping both
objects and arrows from one category to another. Hence a functor has two
components, one which maps objects into objects
$F : \Obj(\CatC) \to \Obj(\CatD)$ and one which maps arrows into arrows
$F : \Hom_{\CatC}(A, B) \to \Hom_{\CatD} (FA,FB)$ such that identity and
composition of arrows are preserved:
\begin{mathpar}
  F(id_A) = id_{FA} \and F(g \circ f) = F(g) \circ F(f)
\end{mathpar}
This latter component is also called the \emph{functorial action} and can be
thought of as the $\operator{fmap}$ program in functional programming.

For example, in $\Set$, we can define the set of lists
$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument,
however, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

@@

\section{Mechanising Extractable Container Functors}
\label{sec:containers}
In order to abstract recursion patters we need to be able to abstract away from
the particular shape of the data.  This is achieved by introducing the concept
of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
who have a initial algebras. A common approach to construct such functors is to
use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:functors}).
However, reasoning about container equality will require us to consider both
functional extensionality and heterogeneous equality. We avoid these axioms by
introducing a custom equivalence relation on types
(Section~\ref{sec:types-morphisms}).

\subsection{Functors and Containers}
\label{sec:functors}
[CUT]
A container is defined by a type of \emph{shapes}
and a \emph{family of position types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
have any type variables and the second constructor has one type variable.

>>>>

%%% TeX-command-extra-options: "-shell-escape"
\documentclass{llncs}
% \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\pagestyle{plain}

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion, in part due to the rigorous set of algebraic laws that
they satisfy. These laws are the basis for reasoning about program equivalences
and, therefore , they can be used for reasoning about program correctness and
optimisations.  Some of these optimisations are successfully applied by
compilers of (functional) languages.  Formalising recursion schemes in a type
theory offers additional termination guarantees, but it often requires
compromises affecting the resulting code, such as imposing performance
penalties, requiring the assumption of additional axioms, or introducing unsafe
casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).

@@

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
reasoning about program equivalences, transformations, and
optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

Despite the rigorous set of algebraic laws that are satisfied by recursion
schemes, there is a lack of tool support for their use in the context of program
calculation. In fact, most of the work on applying program calculation is done
by performing pen-and-paper proofs, and then translating the result to specific
instances of recursion schemes, implemented in a programming language (generally
Haskell). There are some examples of implementations of program calculation
techniques, but these implementations are scarce, not up to date, and not
verified in a proof assistant. For example, Cunha et al.~\cite{DrHylo} automated
program calculation techniques by a custom Haskell implementation.

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the work
in mechanising recursion schemes focuses on a narrow subset of the known
recursion schemes due to termination issues, or do not focus on proving
algebraic laws of generic recursion
schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
most of the mechanisations of generic recursion schemes focus on \emph{maps} and
\emph{folds}, and few authors focus on the \emph{unfolds}. Omitting
\emph{unfolds} severely limits the expressivity of the resulting mechanisations,
and prevent them from being able to mechanise the most general recursion scheme:
a divide-and-conquer algorithm which goes under the name of
\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}. This generality has been proven by
Hinze et al.~\cite{HinzeWG15}, by showing that \emph{all known  recursion
schemes are instances of hylomorphisms}.  To date, no work has mechanised
hylomorphisms in the Coq proof assistant, together with their algebraic laws.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code, and it is not well-suited for
program calculation. This is unfortunate since code extration is what allows the
execution of code that has been verified in
Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
al's approach, extraction (1) does not preserve the recursive structure of
common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
the generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code.
The
full mechanisation is open source.
%can be found on
%Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
%  ANONYMISE THIS}.
While programmers still need to reason about the termination of
their programs, our mechanisation will allow the use of the algebraic laws of
hylomorphisms for \emph{program calculation}, as well as extracting to idiomatic
code (see examples of extracted code in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

To summarise, our contributions are as follows. In this paper we provide a
framework for \emph{program calculation} with recursion schemes in Coq, where
program calculation is done by applying regular Coq equalities, based on  proofs
of the algebraic laws of hylomorphisms, and that allows the extraction of safe,
idiomatic OCaml code:
\begin{itemize}
  \item In Section~\ref{sec:rec-schemes} we give an overview about recursion
schemes.
  \item In Section~\ref{sec:containers} we formalise the type of container
functors ensuring the presence of least and greatest fixed-points for functors
and suitably adapted for program extraction.
  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
hylomorhisms and their universal properties.
  \item In Section~\ref{sec:examples} we use the framework to formalise examples
of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
Furthermore we verify the short-cut fusion optimisation and show the extracted
optimised code to OCaml.
\end{itemize}

\section{Recursion Schemes}\label{sec:rec-schemes}
The structure of data is very similar to the structure of an
algorithm which processes that data. This relationship manifests in the form of
structured recursion schemes which are widely used in functional languages such
Haskell. Canonical examples are folds (catamorphisms), which consume data, and
unfolds (anamorphisms), which produce it. While some implementations of
recursion schemes like $\operator{foldr}$ in Haskell are specific to a particular
data structure, in this case $\operator{List}$s, we can generalise further these
ideas to account for generic (co)inductive data types and generic algorithms
operating on them. This is done by looking at recursion schemes from the point
of view of category theory.

@@

A \emph{functor} $F : \CatC \to \CatD$ is a map between categories mapping both
objects and arrows from one category to another. Hence a functor has two
components, one which maps objects into objects
$F : \Obj(\CatC) \to \Obj(\CatD)$ and one which maps arrows into arrows
$F : \Hom_{\CatC}(A, B) \to \Hom_{\CatD} (FA,FB)$ such that identity and
composition of arrows are preserved:
\begin{mathpar}
  F(id_A) = id_{FA} \and F(g \circ f) = F(g) \circ F(f)
\end{mathpar}
This latter component is also called the \emph{functorial action} and can be
thought of as the $\operator{fmap}$ higher-order function in functional
programming.

For example, in $\Set$, we can define the set of lists
$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.
The terminal object in $\Set$ is the \emph{unit} type, that only has one
inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
The $\List{-}$ type is a functor. Its action on objects is to take any set
$A$ to $\List{A}$, and its functorial
action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

@@

\section{Mechanising Extractable Container Functors}
\label{sec:containers}

In this section, we focus on mechanising \emph{container functors} in a way that
is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
we need to be able toabstract away from the particular shape of the data, and
that this is achieved by introducing the concept of \emph{functors} which have
the suitable fixed-point properties, i.e.\ those which have a initial algebras.
A common approach to construct such functors is to use
\emph{containers}~\cite{AbbottAG05}.  However, reasoning about container
equality will require us to consider both functional extensionality and
heterogeneous equality. We avoid these axioms by introducing a custom
equivalence relation on types (Section~\ref{sec:types-morphisms}).

\subsection{Functors and Containers}
\label{sec:functors}

A straightforward mechanisation of functors as defined in
Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
condition. In fact, if we simply require that functors are functions
\coq{F : Type -> Type}
satisfying the necessary identity and composition properties, then we will not
be able to construct least/greatest fixed points of this type, and we will not
be able to mechanise recursion schemes that follow this structure:
\begin{coqcode}
(* REJECTED due to negative occurrence of (LFix F) *)
Inductive LFix (F : Type -> Type) := LFix_in (lfix_out : F (LFix F)).
\end{coqcode}
One common workaround for this is the use of Mendler-style
recursion~\cite{Mendler91}. Roughly, the idea is to ``pack'' an abstract type
together with the functor that captures the structure of the recursion. From
this abstract type, we will be able to ``extract'' the contents of the functor.
This is, for example, the solution used by the PaCo library~\cite{PaCo}, and a
simplified version of it is the following:
\begin{coqcode}
Inductive MendlerLFix (F : Type -> Type) :=
  In Y (next : Y -> MendlerLFix F) (in_op : F Y).
\end{coqcode}
While this definition solves the strict positivity problem, it does not solve
the problems of: (1) reasoning about data/program equalities; and (2) extraction
to idiomatic OCaml code. To reason about program equalities, we would need to
reason about equalities of functions with \emph{some abstract} \coq{Y} as their
domain.  Furthermore, the extraction of \coq{F Y} will lead to unsafe casts in
the extracted OCaml code.

\paragraph{Containers} Another common solution is to the strict positivity
problem is the use \emph{containers}~\cite{AbbottAG05}.  A container is defined
by a type of \emph{shapes} and a \emph{family of position types} indexed by
shapes.
\begin{coqcode}
(Shape : Type) (Pos : Shape -> Type)
\end{coqcode}

For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
have any type variables and the second constructor has one type variable.
